Next Trippy Structures\thanks{Supported by organization x.}

### 2019-02-21 paper.tex
Outline for experimental section

\newcommand{\acumm}{T-Matrices} % T is for Total!
\newcommand{\ctr}{NewCTR (name subject to change)\ }
\newcommand{\ctrq}{NCTRc}

@@

\begin{figure}[hbt!]
\includegraphics[width=\textwidth]{example_trips.eps}
\caption{An example of five trips represented on \ctr with the optimizations for \texttt{WML} and \texttt{WMJ}, and sections for each stop delimited by dotted lines}
\label{fig:example_ctr}
\end{figure}

@@

 \caption{Extracting the trip \DataSty{i} from \ctr where \FuncSty{CSA}, \FuncSty{WML} and \FuncSty{WMJ} are the structures previously described in (i), (ii) and (iii), respectively}
 \label{alg:extract}
\end{algorithm}

A more complex query that we can solve with \ctr is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window. The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).

@@

    \item In \textbf{line 7} we query \texttt{WML} in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurrences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr was constructed without the optimization that separates \texttt{WML} in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

Of course, \ctr can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, as shown in \cite{brisaboa2018compact}, with the additional possibility of differentiating stops that are only used to switch lines in \ctr. However, there is no known way of using this representation to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).

@@

\begin{threeparttable}
\caption{Efficiency of \acumm~and \ctr against the queries described in Section~\ref{sec:rq}}
\label{tab:queries}
\begin{tabular}{|l|l|l|}
\hline
Query &  \acumm & \ctr\\
\hline
\texttt{board\_X$_{LT}$} & $O(l)$\tnote{a} & \ctrq\tnote{b} \\
\texttt{alight\_X$_{LT}$} & $O(l)$\tnote{a} & Hard\tnote{cd} \\
\texttt{start\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{end\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{switch\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{board\_L$_T$} & O(1) & $O(|S| \times \ctrq)$\tnote{be} \\
\texttt{alight\_L$_T$} & O(1) & $O(|S| \times \ctrq)$\tnote{bce} \\
\texttt{start\_L$_T$} & - & $O(log(|L|\times\bar{|J^l|}))$\tnote{fg} \\
\texttt{end\_L$_T$} & - & $O(|S| \times \ctrq)$\tnote{be} \\
\texttt{board\_T} & $O(|L|)$\tnote{f} & Hard\tnote{d} \\
\texttt{alight\_T} & $O(|L|)$\tnote{f} & Hard\tnote{cd} \\
\texttt{start\_T} & - & $O(|L| \times log(|L|\times\bar{|J^l|}))$\tnote{fg} \\
\texttt{end\_T} & - & Hard\tnote{d} \\
\texttt{load\_X$_{LT}$} & $O(l)$\tnote{a} & Hard\tnote{cd} \\
\hline
\end{tabular}

@@

\begin{itemize}
    \item \texttt{board\_X$_{LT}$} This can be solved easily in the \acumm, as there are aggregated matrices for boarding stops, even though it is necessary to access a separate matrix for every line queried, hence $O(l)$. For \ctr, the number of occurrences for the stop X is counted by delimiting its range on \FuncSty{D} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed. From this number we must subtract the occurrences of final stops, obtained by one end\_X$_{LT}$ with the same restrictions.
    \item \texttt{alight\_X$_{LT}$} In \ctr, the only alighting stops that are explicitly represented are the final stops. To obtain an approximated count of the rest of them, we would need to extract and examine all the occurrences for \texttt{board\_W$_{LT}$} for every W that could have been boarded before X, as the worst-case time complexity of solving with its index operations is prohibitive. On the other hand, it can be solved much faster in \acumm~by accessing the alighting matrices.
    \item \texttt{start\_X$_{LT}$} Can only be solved in \ctr, by delimiting the range for the pattern $\$X$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{end\_X$_{LT}$} Can only be solved in \ctr, by delimiting the range for the pattern $X\$$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{switch\_X$_{LT}$} Calculated as board\_X$_{LT}$ - start\_X$_{LT}$.
    \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} Explained in detail in Algorithm~\ref{alg:xy}.
    \item \texttt{board\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. While in \ctr the total number of occurences of line L can be calculated in one $O(log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    \item \texttt{alight\_L$_T$} Can be approximated in \ctr with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{start\_L$_T$} Can only be solved in \ctr, by filtering down \FuncSty{WML} and \FuncSty{WMJ}, if needed, over the $\$$ section of the \FuncSty{CSA}.
    \item \texttt{end\_L$_T$} Can only be solved in \ctr, by performing a end\_X$_{LT}$ for every X from the line L.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm~and \ctr could solve it by applying board\_L$_T$ $|L|$ times, this would imply a prohibitive cost for \ctr, where extracting all the trips would be preferred in practice.
    \item \texttt{start\_T} Solved in the \ctr by applying \texttt{start\_X$_{LT}$} $|L|$ times.
    \item \texttt{end\_T} Neither of our representations can solve this query efficiently, although it could be approximated by \texttt{start\_T} in CTR.
    \item \texttt{load\_X$_{LT}$} This can be easily calculated in \acumm~as detailed in Section~\ref{sec:acumm}. \marginpar{Tirso please explain how!}
\end{itemize}

\section{Experiments}
\label{sec:exp}
For space, I propose to compare with our previous structures and with a plain representation using bits per symbol (bps). Otherwise, the baseline we select for our $\%$ can be misleading.

@@

For time, we will compare against SPIRE 2018, and maybe, IF WE HAVE TIME, with a plain postgresql baseline. Experiments will be executed in Compostela2 or whatever we'll have available.

>>>>

\newcommand{\acumm}{T-Matrices} % T is for Total!
\newcommand{\ctr}{XCTR}
\newcommand{\ttctr}{TTCTR}
\newcommand{\ctrq}{NCTRc}

@@

\begin{figure}[hbt!]
\includegraphics[width=\textwidth]{example_trips.eps}
\caption{An example of five trips represented on \ctr~with the optimizations for \texttt{WML} and \texttt{WMJ}, and sections for each stop delimited by dotted lines}
\label{fig:example_ctr}
\end{figure}

@@

 \caption{Extracting the trip \DataSty{i} from \ctr, where \FuncSty{CSA}, \FuncSty{WML} and \FuncSty{WMJ} are the structures previously described in (i), (ii) and (iii), respectively}
 \label{alg:extract}
\end{algorithm}

A more complex query that we can solve with \ctr~is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window. The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).

@@

    \item In \textbf{line 7} we query \texttt{WML} in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurrences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr~was constructed without the optimization that separates \texttt{WML} in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

Of course, \ctr~can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, as shown in \cite{brisaboa2018compact}, with the additional possibility of differentiating stops that are only used to switch lines in \ctr. However, there is no known way of using this representation to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).

@@

\begin{threeparttable}
\caption{Efficiency of \acumm~and \ctr~against the queries described in Section~\ref{sec:rq}}
\label{tab:queries}
\begin{tabular}{|l|l|l|}
\hline
Query &  \acumm & \ctr\\
\hline
\texttt{board\_X$_{LT}$} & $O(l)$\tnote{a} & \ctrq\tnote{b} \\
\texttt{alight\_X$_{LT}$} & $O(l)$\tnote{a} & Hard\tnote{cd} \\
\texttt{start\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{end\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{switch\_X$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} & - & \ctrq\tnote{b} \\
\texttt{board\_L$_T$} & O(1) & $O(|S| \times \ctrq)$\tnote{be} \\
\texttt{alight\_L$_T$} & O(1) & $O(|S| \times \ctrq)$\tnote{bce} \\
\texttt{start\_L$_T$} & - & $O(log(|L|\times\bar{|J^l|}))$\tnote{fg} \\
\texttt{end\_L$_T$} & - & $O(|S| \times \ctrq)$\tnote{be} \\
\texttt{board\_T} & $O(|L|)$\tnote{f} & Hard\tnote{d} \\
\texttt{alight\_T} & $O(|L|)$\tnote{f} & Hard\tnote{cd} \\
\texttt{start\_T} & - & $O(|L| \times log(|L|\times\bar{|J^l|}))$\tnote{fg} \\
\texttt{end\_T} & - & Hard\tnote{d} \\
\texttt{load\_X$_{LT}$} & $O(l)$\tnote{a} & Hard\tnote{cd} \\
\hline
\end{tabular}

@@

\begin{itemize}
    \item \texttt{board\_X$_{LT}$} This can be solved easily in the \acumm, as there are aggregated matrices for boarding stops, even though it is necessary to access a separate matrix for every line queried, hence $O(l)$. For \ctr, the number of occurrences for the stop X is counted by delimiting its range on \FuncSty{D} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed. From this number we must subtract the occurrences of final stops, obtained by one end\_X$_{LT}$ with the same restrictions.
    \item \texttt{alight\_X$_{LT}$} In \ctr, the only alighting stops that are explicitly represented are the final stops. To obtain an approximated count of the rest of them, we would need to extract and examine all the occurrences for \texttt{board\_W$_{LT}$} for every W that could have been boarded before X, as the worst-case time complexity of solving with its index operations is prohibitive. On the other hand, it can be solved much faster in \acumm~by accessing the alighting matrices.
    \item \texttt{start\_X$_{LT}$} Can only be solved in \ctr, by delimiting the range for the pattern $\$X$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{end\_X$_{LT}$} Can only be solved in \ctr, by delimiting the range for the pattern $X\$$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{switch\_X$_{LT}$} Calculated as board\_X$_{LT}$ - start\_X$_{LT}$.
    \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} Explained in detail in Algorithm~\ref{alg:xy}.
    \item \texttt{board\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. While in \ctr~the total number of occurences of line L can be calculated in one $O(log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    \item \texttt{alight\_L$_T$} Can be approximated in \ctr~with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{start\_L$_T$} Can only be solved in \ctr, by filtering down \FuncSty{WML} and \FuncSty{WMJ}, if needed, over the $\$$ section of the \FuncSty{CSA}.
    \item \texttt{end\_L$_T$} Can only be solved in \ctr, by performing a end\_X$_{LT}$ for every X from the line L.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm~and \ctr~could solve it by applying board\_L$_T$ $|L|$ times, this would imply a prohibitive cost for \ctr, where extracting all the trips would be preferred in practice.
    \item \texttt{start\_T} Solved in the \ctr~by applying \texttt{start\_X$_{LT}$} $|L|$ times.
    \item \texttt{end\_T} Neither of our representations can solve this query efficiently, although it could be approximated by \texttt{start\_T} in CTR.
    \item \texttt{load\_X$_{LT}$} This can be easily calculated in \acumm~as detailed in Section~\ref{sec:acumm}. \marginpar{Tirso please explain how!}
\end{itemize}

\section{Experiments}
\label{sec:exp}
In this section we discuss the practical performance of our structures. To evaluate them, we have run randomly generated queries against \acumm, \ttctr~and \ctr, with multiple configurations (TODO: describe configurations).

\subsection{Experimental dataset}
Write about the generation of Madrid dataset from real bus journeys.

\subsection{Space requirments}
Missing space tables with total size, compression and bps!

\subsection{Query performance}
Missing a comparison of \acumm vs \ttctr!

@@
